---
title: "R Notebook"
output: html_notebook
---

Packages
===

```{r}
library(ggplot2)
library(dplyr)
library(pitchRx) ##The pitch fx package for R 
##data base management packages
library(DBI)
library(dbplyr)
library(RSQLite)
```

Functions and scraping
===

## How to scrape -- limit of 200 games at a time

```{r}
data(gids, package = "pitchRx")
teamYear <- gids[grepl("mil", gids) & grepl("2011", gids)] ##example is milwaukee 2011, will have to filter to get rid of spring training though

## This function will grab the data for a batter given you have a set of game ids and the batter name
## Inputs: Vector of game ids and a batter name
## Outputs: Data for the batter for all game ids
GetBatterData <- function(teamYear, player) {
  dat <- scrape(game.ids = teamYear)
  locations <- select(dat$pitch, pitch_type, start_speed, px, pz, des, num, gameday_link, inning)
  names <- select(dat$atbat, pitcher, batter, pitcher_name, batter_name, num, gameday_link, event, stand, b, s)
  data <- names %>% filter(batter_name == player) %>% inner_join(locations, ., by = c("num", "gameday_link"))
  return(data)
}

## This function will grab the data for a pitcher given you have a set of game ids and the pitcher name
## Inputs: Vector of game ids and a pitcher name
## Outputs: Data for the pitcher for all game ids
GetPitcherData <- function(teamYear, player) {
  dat <- scrape(game.ids = teamYear)
  locations <- select(dat$pitch, pitch_type, start_speed, px, pz, des, num, gameday_link, inning)
  names <- select(dat$atbat, pitcher, batter, pitcher_name, batter_name, num, gameday_link, event, stand, b, s)
  data <- names %>% filter(pitcher_name == player) %>% inner_join(locations, ., by = c("num", "gameday_link"))
  return(data)
}


brewdata <- GetBatterData(teamYear[-c(1:32)], "Ryan Braun") ##filtered out the spring training games in the teamYear vector
```

## Or how to scrape into a database that has access to all of the games from 2008 - 2015

```{r}
## This whole data base is around 5GB or so, so be careful when downloading
my_db <- src_sqlite("GamedayDB.sqlite3", create = TRUE)
# Collect and store all PITCHf/x data from 2008 to 2015
scrape(start = "2008-01-01", end = "2015-10-04", suffix = "inning/inning_all.xml", connect = my_db$con)
```

Or read in the database if you alread have it

```{r}
my_db <- src_sqlite("GamedayDB.sqlite3")
```

Functions for querying in the data from the database

```{r}
## This function will query the batter data for a certain year for a batter from the data base
## Inputs: batterName, year (as strings)
## Outputs: Data for all of the batter's at bats in the year
QueryBatterData <- function(batterName, year) {
  pitch <- dbGetQuery(my_db$con, paste("SELECT * ",
                                       "FROM pitch ",
                                       "WHERE gameday_link LIKE '%",year,"%'", sep = ""))
  atbat <- dbGetQuery(my_db$con, paste("SELECT * ",
                                       "FROM atbat ",
                                       "WHERE gameday_link LIKE '%",year,"%'", sep = ""))
  bats <- atbat %>% filter(batter_name == batterName)
  data <- bats %>% left_join(pitch, by = c("num", "gameday_link"))
  return(data)
}

## This function will query the pitcher data for a certain year for a pitcher from the data base
## Inputs: pitcherName, year (as strings)
## Outputs: Data for all of the pitcher's pitches during the year
QueryPitcherData <- function(pitcherName, year) {
  pitch <- dbGetQuery(my_db$con, paste("SELECT * ",
                                       "FROM pitch ",
                                       "WHERE gameday_link LIKE '%",year,"%'", sep = ""))
  atbat <- dbGetQuery(my_db$con, paste("SELECT * ",
                                       "FROM atbat ",
                                       "WHERE gameday_link LIKE '%",year,"%'", sep = ""))
  bats <- atbat %>% filter(pitcher_name == pitcherName)
  data <- bats %>% left_join(pitch, by = c("num", "gameday_link"))
  return(data)
}



brewdata <- QueryBatterData("Ryan Braun", "2011")
brewdata <- brewdata[!is.na(brewdata$px),]
```


## Example of getting the last pitch in an at bat

```{r}
braun_last_pitch <- brewdata %>% group_by(gameday_link, num) %>% slice(n()) %>% ungroup()
```

## Function for creating the zones

```{r}
## This function will create the ZONES for the espn style hot and cold zones visual display
## Inputs: A data frame with pitch locations (px and pz)
## Outputs: the same data frame but with the associated zones on each pitch
CreateZones <- function(data) {
  data$Zones <- with(data, ifelse(px > -1 & px < -1/3 & pz > 17/6 & pz < 3.5, 1,
                                  ifelse(px > -1/3 & px < 1/3 & pz > 17/6 & pz < 3.5, 2,
                                         ifelse(px > 1/3 & px < 1 & pz > 17/6 & pz < 3.5, 3,
                                                ifelse(px > -1 & px < -1/3 & pz > 13/6 & pz < 17/6, 4,
                                                       ifelse(px > -1/3 & px < 1/3 & pz > 13/6 & pz < 17/6,5,
                                                              ifelse(px > 1/3 & px < 1 & pz > 13/6 & pz < 17/6,6,
                                                                     ifelse(px > -1 & px < -1/3 & pz > 1.5 & pz < 13/6, 7,
                                                                            ifelse(px > -1/3 & px < 1/3 & pz > 1.5 & pz < 13/6,8,
                                                                                   ifelse(px > 1/3 & px < 1 & pz > 1.5 & pz < 13/6,9,
                                                                                          ifelse(px > -1 & px < -1/3 & pz > 3.5,10,
                                                                                                 ifelse(px > -1/3 & px < 1/3 & pz > 3.5,11,
                                                                                                        ifelse(px > 1/3 & px < 1 & pz > 3.5,12,
                                                                                                               ifelse(px > 1 & pz > 17/6 & pz < 3.5,13,
                                                                                                                      ifelse(px > 1 & pz > 13/6 & pz < 17/6,14,
                                                                                                                             ifelse(px > 1 & pz > 1.5 & pz < 13/6,15,
                                                                                                                                    ifelse(px > 1/3 & px < 1 & pz < 1.5,16,
                                                                                                                                           ifelse(px > -1/3 & px < 1/3 & pz < 1.5,17,
                                                                                                                                                  ifelse(px > -1 & px < -1/3 & pz < 1.5,18,
                                                                                                                                                         ifelse(px < -1 & pz > 1.5 & pz < 13/6,19,
                                                                                                                                                                ifelse(px < -1 & pz > 13/6 & pz < 17/6,20,
                                                                                                                                                                       ifelse(px < -1 & pz > 17/6 & pz < 3.5,21,
                                                                                                                                                                              ifelse(px < -1 & pz > 3.5,22,
                                                                                                                                                                                     ifelse(px > 1 & pz > 3.5,23,
                                                                                                                                                                                            ifelse(px > 1 & pz < 1.5,24,25)))))))))))))))))))))))))
  return(data)
}
```

## Create at bat, hit, on base, plate appearance(all), bases to find avg, obp, ops On the Last_Pitch data set

```{r}
##Classifying what each even means in a statistical setting
hits <- c("Single", "Double", "Triple", "Home Run")
onbase <- c(hits, "Walk", "Intent Walk", "Hit By Pitch")
nonattbat <- c("Sac Fly", "Sac Fly - DP", "Sac Bunt")
##or you can filter by balls in play and do just on batted balls in play
ballsinplay <- c(hits,nonattbat,"Bunt Pop Out", "Bunt Groundout", "Bunt Lineout", "Double Play", "Field Error", "Fielders Choice", "Fielders Choice Out", "Flyout", "Forceout", "Grounded Into DP", "Groundout", "Lineout", "Pop Out", "Triple Play")


## This function will turn the even column into the correct baseball stat
## Inputs: Data frame with the event column
## Outputs: Data frame with added Hit, On Base, At Bat, and number of bases
EventToStat <- function(data) {
  data <- data %>% mutate(Hit = event %in% hits,
                          OnBase = event %in% onbase,
                          AtBat = !(event %in% nonattbat),
                          Bases = ifelse(event == "Single",1,
                                         ifelse(event == "Double",2,
                                                ifelse(event == "Triple",3,
                                                       ifelse(event == "Home Run",4,0)))))
  return(data)
}
```

## Creating the slashline functions

```{r}
## This funtion will compute batting avergae, on base percentage, slugging percentage, and OPS for each hitting zone
## Inputs: Data frame with zones
## Outputs: Data frame with slashlines for each zone
StatComputer <- function(data) {
  dataZones <- data %>% group_by(Zones) %>% summarise(AVG = sum(Hit) / sum(AtBat),
                                                      OBP = sum(OnBase) / n(),
                                                      SLUG = sum(Bases) / sum(AtBat)) %>%
                                                      mutate(OPS = OBP + SLUG)
  return(dataZones)
}

##OR FOR BY COUNT

## This function will compute the slash lines for each hitting zone per count
## Inputs: Data frame with zones
## Outputs: Data frame with stats per zone per count
StatsByCounts <- function(data) {
  data$b[data$b == 4] <- 3
  data$s[data$s == 3] <- 2
  dataZones2 <- data %>% group_by(Zones,b,s) %>% summarise(AVG = sum(Hit) / sum(AtBat),
                                                      OBP = sum(OnBase) / n(),
                                                      SLUG = sum(Bases) / sum(AtBat)) %>%
                                                      mutate(OPS = OBP + SLUG)
  return(dataZones2)
}
```

## Making the grid set up for the visualizations

```{r}
## This function will make the points to plot the visualization for the zones
## Inputs: zones data frame
## Outputs: zones data frame with points
MakeGrid <- function(data) {
  data <- data[!is.na(data$Zones),] ##Get rid of the NA group

  data$X <- c(rep(-2/3,nrow(data[data$Zones == 1,])),rep(0,nrow(data[data$Zones == 2,])),rep(2/3,nrow(data[data$Zones == 3,])),rep(-2/3,nrow(data[data$Zones == 4,])),rep(0,nrow(data[data$Zones == 5,])),rep(2/3,nrow(data[data$Zones == 6,])),rep(-2/3,nrow(data[data$Zones == 7,])),rep(0,nrow(data[data$Zones == 8,])),rep(2/3,nrow(data[data$Zones == 9,])),rep(-2/3,nrow(data[data$Zones == 10,])),rep(0,nrow(data[data$Zones == 11,])),rep(2/3,nrow(data[data$Zones == 12,])),rep(4/3,nrow(data[data$Zones == 13,])),rep(4/3,nrow(data[data$Zones == 14,])),rep(4/3,nrow(data[data$Zones == 15,])),rep(2/3,nrow(data[data$Zones == 16,])),rep(0,nrow(data[data$Zones == 17,])),rep(-2/3,nrow(data[data$Zones == 18,])),rep(-4/3,nrow(data[data$Zones == 19,])),rep(-4/3,nrow(data[data$Zones == 20,])),rep(-4/3,nrow(data[data$Zones == 21,])),rep(-4/3,nrow(data[data$Zones == 22,])),rep(4/3,nrow(data[data$Zones == 23,])),rep(4/3,nrow(data[data$Zones == 24,])),rep(-4/3,nrow(data[data$Zones == 25,])))

  data$Z <- c(c(rep(19/6,nrow(data[data$Zones == 1,])),rep(19/6,nrow(data[data$Zones == 2,])),rep(19/6,nrow(data[data$Zones == 3,])),rep(15/6,nrow(data[data$Zones == 4,])),rep(15/6,nrow(data[data$Zones == 5,])),rep(15/6,nrow(data[data$Zones == 6,])),rep(11/6,nrow(data[data$Zones == 7,])),rep(11/6,nrow(data[data$Zones == 8,])),rep(11/6,nrow(data[data$Zones == 9,])),rep(23/6,nrow(data[data$Zones == 10,])),rep(23/6,nrow(data[data$Zones == 11,])),rep(23/6,nrow(data[data$Zones == 12,])),rep(19/6,nrow(data[data$Zones == 13,])),rep(15/6,nrow(data[data$Zones == 14,])),rep(11/6,nrow(data[data$Zones == 15,])),rep(7/6,nrow(data[data$Zones == 16,])),rep(7/6,nrow(data[data$Zones == 17,])),rep(7/6,nrow(data[data$Zones == 18,])),rep(11/6,nrow(data[data$Zones == 19,])),rep(15/6,nrow(data[data$Zones == 20,])),rep(19/6,nrow(data[data$Zones == 21,])),rep(23/6,nrow(data[data$Zones == 22,])),rep(23/6,nrow(data[data$Zones == 23,])),rep(7/6,nrow(data[data$Zones == 24,])),rep(7/6,nrow(data[data$Zones == 25,]))))
return(data)
}
```

Setting up the data to plot
===

```{r}
##EVERYTHING
braun <- CreateZones(braun_last_pitch)
braun <- EventToStat(braun)
braun <- StatComputer(braun)
braun <- MakeGrid(braun)
##OR BY COUNT
braun2 <- CreateZones(braun_last_pitch)
braun2 <- EventToStat(braun2)
braun2 <- StatsByCounts(braun2)
braun2 <- MakeGrid(braun2)
##OR JUST BALLS IN PLAY
braun3 <- CreateZones(braun_last_pitch)
braun3 <- EventToStat(braun3 %>% filter(event %in% ballsinplay))
braun3 <- StatComputer(braun3)
braun3 <- MakeGrid(braun3)
```

Plotting
===

For all plots you can plot by AVG, OBP, SLUGG, or OPS

```{r}
ggplot(data = braun, aes(x = X, y = Z, col = AVG)) +
  geom_point(size = 20, shape = 15) +
  geom_label(aes(x = X, y = Z, label = round(AVG,3))) +
  theme_bw() +
  scale_x_continuous(limits = c(-3.5,3.5)) +
  scale_y_continuous(limits = c(0.5,4.5)) +
  scale_color_gradient(low = "blue", high = "red") +
  annotate("rect", xmin = -1, xmax = 1, ymin = 1.5, ymax = 3.5, color = "black", alpha = 0, size = 2) +
  ylab("Horizontal Location (ft.)") +
  xlab("Vertical Location (ft): Catcher's View") +
  labs(color = "AVG") +
  ggtitle("Ryan Braun 2011: Batting Avergage Per Pitch Location")
```

BY COUNTS

```{r}
ggplot(data = braun2, aes(x = X, y = Z, col = AVG)) +
  geom_point(size = 5, shape = 15) +
  facet_wrap(b~s) +
  theme_bw() +
  scale_x_continuous(limits = c(-3.5,3.5)) +
  scale_y_continuous(limits = c(0.5,4.5)) +
  scale_color_gradient(low = "blue", high = "red") +
  annotate("rect", xmin = -1, xmax = 1, ymin = 1.5, ymax = 3.5, color = "black", alpha = 0, size = 2) +
  ylab("Horizontal Location (ft.)") +
  xlab("Vertical Location (ft): Catcher's View") +
  labs(color = "AVG") +
  ggtitle("Ryan Braun 2011: Batting Avergage Per Pitch Location by Count")
```

BATTED BALLS IN PLAY

```{r}
ggplot(data = braun3, aes(x = X, y = Z, col = AVG)) +
  geom_point(size = 20, shape = 15) +
  geom_label(aes(x = X, y = Z, label = round(AVG,3))) +
  theme_bw() +
  scale_x_continuous(limits = c(-3.5,3.5)) +
  scale_y_continuous(limits = c(0.5,4.5)) +
  scale_color_gradient(low = "blue", high = "red") +
  annotate("rect", xmin = -1, xmax = 1, ymin = 1.5, ymax = 3.5, color = "black", alpha = 0, size = 2) +
  ylab("Horizontal Location (ft.)") +
  xlab("Vertical Location (ft): Catcher's View") +
  labs(color = "AVG") +
  ggtitle("Ryan Braun 2011: Batting Avergage on Batted Balls in Play per Pitch Location")
```

Other fun plots
===

All of Braun's pitches seen in 2011

```{r}
ggplot(data = brewdata, aes(x = px, y = pz, col = start_speed)) +
  geom_point(size = 4, alpha = 0.05) +
  theme_bw() +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(0,5)) +
  annotate("rect", xmin = -1, xmax = 1, ymin = 1.5, ymax = 3.5, color = "black", alpha = 0) +
  ylab("Horizontal Location (ft.)") +
  xlab("Vertical Location (ft): Catcher's View") +
  labs(color = "Velocity (mph)", shape = "Pitch Type") +
  ggtitle("Ryan Braun 2011")
```

